---
title: "Access Wideband Audiology Immitance database using R and dplyr"
author: "Nicholas Horton (nhorton@amherst.edu) and Ben Baumer"
date: "November 14, 2017"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---


```{r, include=FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
require(mosaic)
```

```{r, include=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).

# This changes the default colors in lattice plots.
trellis.par.set(theme=theme.mosaic())  

# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small"    # slightly smaller font for code
)
```

## Introduction

This document is intended to describe how to access data from a MySQL database using R.  It utilizes a database of wideband acoustic immitance variables from humans with normal hearing (see https://projectreporter.nih.gov/project_info_description.cfm?aid=8769352&icde=30039221&ddparam=&ddvalue=&ddsub=&cr=10&csb=default&cs=ASC for more details).

A relevant paper on the topic of data management and databases in R can be found at http://chance.amstat.org/2015/04/setting-the-stage.


## Accessing data from a database using SQL commands

First we demonstrate how to access data using SQL (structured query language) commands and the `dbGetQuery()` function.  We begin by setting up a connection to the database.

```{r}
library(mosaic)
library(RMySQL)
con <- dbConnect(MySQL(), host = "scidb.smith.edu",
                user = "waiuser", password = "smith_waiDB", 
                dbname = "wai")
```

Next a series of SQL queries can be sent to the database.  These return R dataframes.

```{r}
dbGetQuery(con, "SHOW TABLES")
dbGetQuery(con, "EXPLAIN PI_Info")
ds <- dbGetQuery(con, "SELECT * from Measurements LIMIT 10")
ds
```

## Accessing a database using dplyr commands

Alternatively, a connection can be made to the server by creating a series of dplyr table objects.  
```{r}
library(mosaic)      
db <- src_mysql(dbname = "wai", host = "scidb.smith.edu", user = "waiuser", 
          password="smith_waiDB")
Measurements <- tbl(db, "Measurements")
PI_Info <- tbl(db, "PI_Info")
Subject <- tbl(db, "Subject")
```

#### Let's explore the `PI_Info` table.
```{r}
PI_Info  %>% summarise(total = n())
PI_Info %>% collect() %>% data.frame()   # collect() is a bad idea when dealing with large tables!
```

#### Let's explore the `Subjects` table.
```{r}
Subject  %>% summarise(total = n())
Subject %>% collect()  # be careful with collect() with large tables!
```

#### Let's explore the `Measurements` table.
```{r}
Measurements %>% summarise(total = n())
```

#### Let's download the data from a given subject
```{r}
onesubj <- 
  Measurements %>% 
  filter(Identifier=="Rosowski_2012", Sub_Number==3) %>%
  collect %>%
  mutate(SessionNum = as.factor(Session))
head(onesubj)
```

#### Finally we can plot the results
```{r eval=TRUE, fig.height=6}
onesubj <- mutate(onesubj, 
  Ear=ifelse(Left_Ear==1, "Left", "Right"))
ggplot(data = onesubj, aes(x = Freq, y = Absorbance)) + geom_point() +
  aes(colour = Ear) + scale_x_log10() + labs(title="Absorbance by ear Rosowski subject 3")
```
